---
title: "Optimization, and review"
author: "Peter Ralph"
date: "19 February 2018 -- Advanced Biological Statistics"
---

```{r setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,
                      fig.height=fig.dim,
                      fig.align='center')
set.seed(23)
library(lars)
library(tidyverse)
library(rstan)
library(matrixStats)
options(mc.cores = parallel::detectCores())
```

# Overview

## Building models

So far we've focused on *building models* for the data.

. . .

Models involve *parameters*, that are often the target of our inference.

. . .

We put *priors* on parameters, for several reasons:

1. To be able to communicate uncertainty using the *posterior*.

2. To incorporate prior information.

3. To "strongly encourage" certain model requirements (e.g., sparsity).


## Some remaining topics: branching out

1. Dimension reduction and visualization (e.g., PCA)

2. Clustering and categorization

3. Time series

4. Spatial and network models

. . .

These all involve new *models*
and mew ways of using *priors*
to achieve analysis goals.


# Optimization

## Another trick up Stan's sleeve

In addition to sampling from the posterior distribution,
Stan can do *optimization*:
hill climb to the top.

. . .

**Definition:** the *maximum a posteriori* (MAP) estimate
is the set of parameter values that maximize the posterior likelihood.

. . .

Recall that
$$\begin{aligned}
    \text{posterior} = \text{prior} \times \text{likelihood} .
\end{aligned}$$
... so this is closely related to the *maximum likelihood* estimate (MLE).

---------------------

```
optimizing                package:rstan                R Documentation

Obtain a point estimate by maximizing the joint posterior

Description:

     Obtain a point estimate by maximizing the joint posterior from the
     model defined by class ‘stanmodel’.

Usage:

     ## S4 method for signature 'stanmodel'
     optimizing(object, data = list(), 
         seed = sample.int(.Machine$integer.max, 1), init = 'random', 
         check_data = TRUE, sample_file = NULL, 
         algorithm = c("LBFGS", "BFGS", "Newton"),
         verbose = FALSE, hessian = FALSE, as_vector = TRUE, 
         draws = 0, constrained = TRUE, ...)   
     
Arguments:

  object: An object of class ‘stanmodel’.

    data: A named ‘list’ or ‘environment’ providing the data for the
          model or a character vector for all the names of objects used
          as data.  See the Note section in ‘stan’.

```

## How to do it

::: {.columns}
::::::::: {.column width=50%}

```r
pois_block <- "
data {
    int N; // number of obs
    int Z[N]; // counts
}
parameters {
    real<lower=0> lambda;
}
model {
    Z ~ poisson(lambda);
}
"
pois_model <- stan_model(
                model_code=pois_block)

Z <- rpois(20, 5)
pois_opt <- optimizing(pois_model,
                       data=list(N=20,
                                 Z=Z))
```



:::
::::::::: {.column width=50%}

```{r pois_opt, echo=FALSE, cache=TRUE}
pois_block <- "
data {
    int N; // number of obs
    int Z[N]; // counts
}
parameters {
    real<lower=0> lambda;
}
model {
    Z ~ poisson(lambda);
}
"
pois_model <- stan_model(model_code=pois_block)

Z <- rpois(20, 5)
(pois_opt <- optimizing(pois_model,
                        data=list(N=20,
                                  Z=Z)))
```


:::
:::::::::


## It is fast

::: {.columns}
::::::::: {.column width=50%}


```{r pois_timing, cache=TRUE}
timings <- lapply(10^(1:5), 
    function (N) {
        Z <- rpois(N, 5)
        a <- system.time(
                 optimizing(pois_model,
                            data=list(N=N,
                                      Z=Z)))
        b <- system.time(
                 stan(model_code=pois_block,
                      data=list(N=N,
                                Z=Z)))
        list(optim=a, mcmc=b) } )
```

:::
::::::::: {.column width=50%}

```{r plot_pois_stan, fig.width=1.5*fig.dim, fig.height=1.5*fig.dim, echo=FALSE}
matplot(10^(1:5), 
        do.call(rbind, lapply(timings, sapply, "[[", "elapsed")),
        type='l', lty=1, log='x',
        xlab="number of points", ylab="seconds, elapsed")
```

:::
:::::::::

-----------------------

The downside of *point estimates* is

. . .

that you've got no estimate of *uncertainty*.


## Another shortcut: "variational Bayes"


```{r pois_vb, echo=FALSE, cache=TRUE}
Z <- rpois(1e5, 5)
(pois_vb <- vb(pois_model,
               data=list(N=1e5,
                         Z=Z)))
```

------------------

```{r show_vb}
pois_vb
```



# Exercises

## Write models, optimize

Let's practice *writing models*,
and compare the resuls of `stan( )` to `optimizing( )`.

----------------------

1. Write down a model on the whiteboard.

2. Explain the model, how to find what you want from it,
   to another pair of people.

3. Code up the Stan model.

4. Simulate some test data.

5. Run `optimizing( )` to get point estimates.


## Pick a situation

1. Number of mosquitos caught in traps
   at 20 different time points at 4 locations;
   temperature and rainfall are also measured.

2. Transpiration rates of 5 trees each of 100 strains,
   along with genotype at five SNPs putatively linked to stomatal efficiency.

3. Presence or absence of *Wolbachia* parasites
   in fifty flies are sampled from each of 100 populations,
   along with the sex and transcription levels of ten immune-related genes of each fly.

*Modifications:* 
(a) change the numbers - 1,000 SNPs instead of five?
(b) make it robust (to outliers)!
