---
title: "beta-binom_analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Question:** We flip an odd-looking coin 100 times, and get 65 heads. What is it’s true* probability of heads?

## What is True?

The long-run proportion of heads (if we flipped it a lot).

## What prior to use?

We'll use $\alpha = \beta = 1$ : a flat prior.

## Analysis:

We know that the posterior distribution on $P$, the true probability of heads,
is Beta($\alpha + z$, $\beta + n - z$).
Here $z=65$ and $n=100$.
We'll make a plot of the density function.

```{r bayes}
z <- 65
n <- 100
prior_alpha <- 1
prior_beta <- 1
post_alpha <- prior_alpha + z
post_beta <- prior_beta + n - z
x <- seq(0, 1, length.out=1000)
dens <- dbeta(x, shape1=post_alpha, shape2=post_beta)
plot(x, dens, type='l',
     xlab='prob of heads', ylab='posterior density')
```

## Is it reasonable that $\theta=1/2$?

It looks like not: the posterior probability
that $P \le 1/2$ is `r pbeta(1/2, shape1=post_alpha, shape2=post_beta)`.

## Best guess at $\theta$?

We'll take the maximum a posteriori estimate (MAP), which is
`r (post_alpha - 1)/(post_alpha + post_beta - 2)`.

## How far off are we, probably?

How much does the answer depend on the prior?

Does our procedure work on simulated data?